TIMES Ireland Model: Residential Sector

### 2021-05-24 manuscript.tex
24May2021

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports. Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for these barriers, but Ireland still has ambitious national climate targets. To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions. Ireland has ambitious targets, despite distinct barriers. The Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets.

@@

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
A Energy System Optimisation Model (ESOM) can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex economic sectoral interconnections within each time period. The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA). TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
\par In addition to TIMES there are other ESOM frameworks available such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. \par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail. Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}. There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets. Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.

@@

The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions. Fig. \ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity. Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.

@@

@@

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.  
In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

The 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}. BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.

Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.

\subsection{Residential Energy Services}
\label{Intro:EnergyServices}

Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage. Other residential energy services and their share of final energy in 2018 is shown in \ref{fig:Ireland2018ServiceDemands}. As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline. The new A-rated dwellings can be used to predict the share of future energy service demands. From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling. However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household. Cooking and Appliances energy usage would be expected to remain stable.

\begin{figure}[!htbp]
\centering
    \begin{tikzpicture}
    \tikzstyle{every node}=[font={\bf}]
    \pie[   color = { yellow!90!black, green!60!black, blue!60,
        red!70,
        gray!80,
        teal!20
    }, explode = 0.4, text=legend, radius=3]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
    \end{tikzpicture}
    \caption{Ireland 2018 Energy Service Demands}
    \label{fig:Ireland2018ServiceDemands}
\end{figure}

Notably space cooling is absent from Fig \ref{fig:Ireland2018ServiceDemands}, but space cooling is accounted for in appliances. The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant increase in space cooling energy is expected in the Ireland residential sector.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    x tick label style={
		/pgf/number format/1000 sep=},xlabel={Year},
	y tick label style={
	    /pgf/number},
		ylabel={Energy Consumption($ktoe$)},
		 scaled y ticks = false,
    xmin=2000, xmax=2050,
    ymin=0, ymax=40,
    xtick={2000,2010,2020,2030,2040,2050},
    ytick={0,10,20,30,40},
    ymajorgrids=true,
]
\addplot[
    color=blue,
    mark=square,
    ]
    coordinates {
    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
    };
\end{axis}
\end{tikzpicture}
    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
    \label{fig:Cooling}
\end{figure}

The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector. Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER label and age of dwelling.

@@

The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture. Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.

Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}. Some of these actions included:
\begin{itemize}
    \item{Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings}
    \item{500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent}
    \item{Installation of 600,000 heat pumps, with 400,000 of these in existing buildings}
    \item{Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes}
    \item{Complete the roll-out of the Support Scheme for Renewable Heat}
    \item{Increasing the number of Sustainable Energy Communities to 1,500}
\end{itemize}

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=1.0]{Figures/CAP2019MACC.png} 
 \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
 \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{figure}

\par
The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$. over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. The options specific to the residential sector are: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{itemize}

Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie. save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie. cost money) but the expected abatement of carbon is far greater.

\subsubsection{Electricity}
\par
The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
Decarbonising electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
In Ireland, a €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce fossil fuel demand and investment. The carbon tax receipts grew every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. Many countries have taken this approach, especially in Europe, where Sweden were second to introduce a carbon tax after Finland, the Swedish carbon tax rate is the highest in the world at €108.81/ $tCO_2$.  About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}. 
The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period. The last REFIT schemes closed in December 2015. In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. RESS delivers for communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW by 2030.

\subsubsection{Heat}
%Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
Sustainable Energy Authority of Ireland (SEAI) have been played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes. SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process. In parallel, new dwellings must achieve a minimum of 20\% renewable energy, the ambient heat used in heat pumps is included as a renewable energy, often heat pumps are the preferred heating technology despite the high capital costs. As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.  The energy efficiency first approach in how Ireland aims to decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
\end{itemize}

\subsubsection{Gas \& District Heating}

In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019AIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and about 13\% Hydrogen\cite{Ervia2019AIreland}. The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future. Increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.

>>>>

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports. Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for these barriers, but Ireland still has ambitious national climate targets. To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions. It can be observed that Ireland has ambitious targets, despite distinct barriers.

@@

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
The Climate Change Advisory Council (CCAC) will rely a range of modelling tools to provide insights to the Irish policymakers. One such model, is an Energy System Optimisation Model (ESOM), which can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex interconnections within each time period. The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA). TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
\par In addition to TIMES there are other ESOM frameworks available such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. \par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail. Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}. There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets. Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.

@@

The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Under these guidelines, Ireland’s residential sector accounted for 14.2\% of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Focusing on energy-related GHG emissions only, the residential sector directly accounts for 24\% of GHG emissions. Fig. \ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity.

@@

Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in decarbonising energy in Ireland.

@@

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Residential GHG emissions are still growing as preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.
In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

The 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
\subsection{Residential Energy Services}
\label{Intro:EnergyServices}

Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage in 2018. Other residential energy services and their share of final energy is shown in Fig.\ref{fig:Ireland2018ServiceDemands}. As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline. The new A-rated dwellings can be used as a guide to show the share of future energy service demands. From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling. However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household. Cooking and Appliances energy usage would be expected to remain stable.

\begin{figure}[!htbp]
\centering
    \begin{tikzpicture}
    \tikzstyle{every node}=[font={\bf}]
    \pie[   color = { yellow!90!black, green!60!black, blue!60,
        red!70,
        gray!80,
        teal!20
    }, explode = 0.6, text=legend, radius=2.4]{61.8/Space Heating, 2.2/Cooking, 18.5/Water Heating, 3.8/Lighting, 12.8/Appliances,0.9/Pumps \& Fans}
    \end{tikzpicture}
    \caption{Ireland 2018 Energy Service Demands}
    \label{fig:Ireland2018ServiceDemands}
\end{figure}

Notably absent from Fig.\ref{fig:Ireland2018ServiceDemands}, is space cooling, which is accounted for in appliances. The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output expected space cooling energy consumption projections are shown in Fig.\ref{fig:Cooling}.

\begin{figure}[!htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    x tick label style={
		/pgf/number format/1000 sep=},xlabel={Year},
	y tick label style={
	    /pgf/number},
		ylabel={Energy Consumption($ktoe$)},
		 scaled y ticks = false,
    xmin=2000, xmax=2050,
    ymin=0, ymax=40,
    xtick={2000,2010,2020,2030,2040,2050},
    ytick={0,10,20,30,40},
    ymajorgrids=true,
]
\addplot[
    color=blue,
    mark=square,
    ]
    coordinates {
    (2000,0.6)(2010,3.1)(2020,8.4)(2030,16.3)(2040,26.1)(2050,37)
    };
\end{axis}
\end{tikzpicture}
    \caption{Space Cooling Residential Energy Consumption (Source: POTEnCIA) }
    \label{fig:Cooling}
\end{figure}

The SEAI residential energy report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} provides high resolution detail on energy demands in the residential sector. Variables such as disposable income, appliance ownership, dwelling location, type of occupier status (owner/rented ), BER labels and age of dwelling.

@@

The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.  The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.

Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}. The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities.

The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$ over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. The options specific to the residential sector are: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{itemize}

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=1]{Figures/CAP2019MACC.png} 
 \caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
 \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{figure}
Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie. save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie. cost money) but the expected abatement of carbon is far greater.

\subsubsection{Electricity}
\par
In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
Decarbonising electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
In Ireland, a €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce fossil fuel demand and investment. The carbon tax receipts grew every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. Many countries have taken this approach, especially in Europe, where Sweden were second to introduce a carbon tax after Finland, the Swedish carbon tax rate is the highest in the world at €108.81/ $tCO_2$.  About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}. 
\par The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period. The last REFIT schemes closed in December 2015. In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. RESS prioritizes communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW solar by 2030.

\subsubsection{Heat}
Sustainable Energy Authority of Ireland (SEAI) have played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes. SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process. In parallel, new dwellings must achieve a minimum of 20\% renewable energy. The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs. As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, shows Irelands energy efficiency first approach in the residential sector.  in how Ireland aims to decarbonize the residential sector. Domestic supports for renewable heating technologies and energy efficiency include:
\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
    \item{ Solar Photovoltaics }
\end{itemize}

\subsubsection{Gas}

In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}, with lower indigenous gas expected GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and 13\% Hydrogen\cite{Ervia2019VisionIreland}. The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future. ALthough increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.
